---
title: "Raster Deepdive"
author: "Nicola Noger"
format: html
editor: visual
engine: knitr
---

# Solution for Raster Deepdive

In this document, I solve the tasks for *Raster Deepdive* of the course *Spatiotemporal Datascience*.

## Load libraries and data

```{r}
library(terra)
library(dplyr)
library(tictoc)
library(readr)

# Ensure knitr can find Git Bash on Windows
bash_path <- Sys.which("bash")
if (bash_path == "") {
  bash_candidates <- c(
    "C:/Program Files/Git/usr/bin/bash.exe",
    "C:/Program Files/Git/bin/bash.exe"
  )
  bash_path <- bash_candidates[file.exists(bash_candidates)][1]
}
if (!is.na(bash_path) && bash_path != "") {
  knitr::opts_chunk$set(engine.path = list(bash = bash_path))
}

# Start runtime measurement
tic("Task 1 (Raster Approach) runtime")

data_dir <- "A:/STUDIUM/06_Fruelingssemester26/STDS/Week_3/data"

# Read vector data
kantone <- vect(file.path(data_dir, "swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"), layer = "tlm_kantonsgebiet")
wald <- vect(file.path(data_dir,"SWISSTLM3D_2026_LV95_LN02.gpkg"), layer = "tlm_bb_bodenbedeckung")

# Keep forest classes
wald <- wald[wald$objektart %in% c("Wald", "Wald offen"), ]


## Task 1: Raster Approach

# Create a template raster based on the extent of the cantons
r_template <- rast(kantone, resolution = 100)

# Rasterize cantons: assign the canton name to each pixel
r_kantone <- rasterize(kantone, r_template, field = "name")

# Rasterize forest: assign 1 to forest pixels, 0 to background
r_wald <- rasterize(wald, r_template, field = 1, background = 0)
names(r_wald) <- "wald_pixel_count"

# Count forest pixels per canton using zonal statistics
wald_pro_kanton <- zonal(r_wald, r_kantone, fun = "sum", na.rm = TRUE)

# Calculate forest area in square meters (1 pixel = 100m * 100m = 10,000 m^2)
wald_pro_kanton <- wald_pro_kanton |> 
  mutate(wald_area = wald_pixel_count * 10000)

# Calculate total canton area (using the fast vector method) and join
kantone$kanton_area <- expanse(kantone)

kanton_frac <- kantone |>
  as.data.frame() |> 
  group_by(name) |>
  summarise(kanton_area = sum(kanton_area)) |>
  left_join(wald_pro_kanton, by = "name") |> 
  mutate(wald_frac = as.numeric(wald_area / kanton_area))

# End and log runtime
terra_time <- tictoc::toc(quiet = TRUE, log = TRUE)
terra_time <- as.numeric(terra_time$toc - terra_time$tic)
```

## Forest share by canton (descending)

```{r}
kanton_frac_sorted <- kanton_frac |>
  arrange(desc(wald_frac)) |>
  select(name, wald_frac) |>
  mutate(wald_frac_pct = round(wald_frac * 100, 2))

knitr::kable(kanton_frac_sorted, 
             col.names = c("Canton", "Forest share", "Forest share (%)"),
             digits = 4)
```


```{r}
tictoc::tic(msg = "Task 2: GDAL - rasterize")
```

```{bash}

# Set file paths
# 1. Define paths
data_dir="A:/STUDIUM/06_Fruelingssemester26/STDS/Week_3/data"
out_dir="A:/STUDIUM/06_Fruelingssemester26/STDS/Week_3/Raster_deepdive/output"
kantone_gpkg="${data_dir}/swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"
wald_gpkg="${data_dir}/SWISSTLM3D_2026_LV95_LN02.gpkg"

# Ensure Git Bash coreutils and conda are available
export PATH="/usr/bin:/bin:$PATH"
CONDA="/c/Users/Nicola/anaconda3/Scripts/conda.exe"
if [ ! -x "$CONDA" ]; then
  CONDA="/c/Users/Nicola/Anaconda3/Scripts/conda.exe"
fi

# 2. Read extent
extent=$($CONDA run -n geo_gdal ogrinfo -so "${kantone_gpkg}" tlm_kantonsgebiet | grep Extent | tr -d '(),-' | awk '{print $2, $3, $4, $5}')

# 3. Rasterize forest (force NoData=255 so 0 counts as non-forest)
$CONDA run -n geo_gdal gdal_rasterize -l tlm_bb_bodenbedeckung -burn 1 -init 0 -a_nodata 255 -ot Byte -tr 50 50 -tap -te ${extent} -where "objektart IN ('Wald', 'Wald offen')" "${wald_gpkg}" "${out_dir}/forest.tif"

# 4. GDAL pipeline (counts all pixels per canton)
$CONDA run -n geo_gdal gdal pipeline read "${out_dir}/forest.tif" ! zonal-stats --zones "${kantone_gpkg}" --zones-layer tlm_kantonsgebiet --stat count --stat sum --include-field name ! write --overwrite --output-format CSV "${out_dir}/zonal_stats.csv"
```

```{r}
gdal_time <- tictoc::toc(quiet = TRUE, log = TRUE)

```

```{r}
gdal_results <- read_csv("A:/STUDIUM/06_Fruelingssemester26/STDS/Week_3/Raster_deepdive/output/zonal_stats.csv")

# Calculate forest share (sum / count)
kanton_frac_gdal_sorted <- gdal_results |>
  mutate(wald_frac = sum / count) |>
  arrange(desc(wald_frac)) |>
  select(name, wald_frac) |>
  mutate(wald_frac_pct = round(wald_frac * 100, 2))
# Show results
knitr::kable(kanton_frac_gdal_sorted, 
             col.names = c("Canton", "Forest share", "Forest share (%)"),
             digits = 4)
```


## Runtime comparison

```{r}

terra_sec <- as.numeric(terra_time)

gdal_sec <- as.numeric(gdal_time$toc - gdal_time$tic)

timings <- data.frame(
  Method = c("sf (Vector Deepdive)", "DuckDB (Vector Deepdive)", "Terra (Raster Task 1)", "GDAL (Task 2)"),
  Time_seconds = c(208.30, 123.14, terra_sec, gdal_sec)
)

knitr::kable(timings, digits = 2)
```

## Results

The raster approaches with `terra` and GDAL produce the same forest shares as the vector computations (sf/DuckDB). The canton order and percentages are consistent, so the results are methodologically identical.

## Discussion

Raster processing was much faster than the vector calculations because the raster data are smaller here and the evaluation is essentially a simple pixel count. The repeated OOM issues with `terra` were unexpected and even crashed my device, which suggests high memory demand during rasterization or intermediate steps despite the smaller data footprint. Overall, the comparison confirms that raster is more efficient here, but `terra` was unstable in my setup.

## Use of AI

AI tools (GitHub Copilot) were used in this exercise to assist with writing and debugging R code, as well as for proofreading and improving text.